Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-08-15 numerics_paper/version2/Response_to_Reviewers.tex

\bigskip
{\color{red}
  \begin{itemize}
  \item[(2)]
    Assuming there is a solid reason for the proposed fix in the paper. It is also unclear how the indicator
    ($\theta$, after Eq. 19b) works to actively switch between Rusanov to Roe and vice-versa. The authors
    mention that $\theta=0$ means Roe-flux whereas $\theta=1$ recovers the Rusanov flux. What is the guiding
    principle to select the 'appropriate' $\theta$ values? This is somewhat discussed in 3.1 in between Eq. 
    16-19b but it is not intuitively clear to me how the entropy-residual is used as a guide to select $\theta$.
    Moreover, what does the value in between $0<\theta<1$ means?
  \end{itemize}
}
\noindent
We provide a precise definition of $\theta_i$ in (19) and (21). The guiding principle to select $\theta_i$
is based on the entropy-residual. As discussed before, using the entropy-residual to locate shocks has proven
its worth in many works. 
By selecting $\theta_i=\{0,1\}$, the numerical flux (15) is either Roe's or Rusanov's.
Having $0<\theta<1$ produces a solver that combines the dissipation of both solvers; i.e.,
for those cases, the blended solver is more dissipative than Roe's but not as much as Rusanov's. 
We have revised the text in Section 3.1 accordingly.

\bigskip
{\color{red}
  \begin{itemize}
  \item[(3)]
    If we continue to assume that (2) above can also be justified, I strongly suggest that the one or two of
    the test cases to be further evaluated. Dumbser et al. (04) and Kitamura et al. (AIAA 09) showed that the
    instability may manifest itself for a certain 'disturbance' in the test cases either via initial point values
    disturbance or grid disturbances, even if the fix works nicely on the non-disturbed conditions. I strongly
    suggest to vary the mesh as done in Kitamura et al. for the cylinder case (at least 3 different meshes for
    cylinder) and to include the residual plot. This is because the residual plot for convergence can easily show
    potential instability. The contour plot solution may look nice but the residual plot may show that it is
    moving toward instability.
  \end{itemize}
}
\noindent
Our initial condition is given by $h_0=1$, $u_0=5$ and $v_0=0$; i.e., we do not start with a shock. 
At the beginning of the simulation, the shock forms near the cylinder and travels to its position at 
steady state. Once at steady state, the shock is not aligned with a grid interface. Instead, the 
shock is resolved within a few cells (see Figure 3c). 
Since we start from a condition far from the steady state, 
perturbing the initital condition or the grid has no effect in the solution. We modified the text in
Section 4.3 to explain this. In addition, we added a plot of the residual; see equation (29) and 
Figure 4. 
The residual stagnates when we use the blended solver with the second-order method (6); however, it 
continues to decrease if the first-order scheme (7) is considered. This suggests that the residual stagnates 
not due to the Riemann solver but due to the second-order corrections.

\bigskip
{\color{red}
  \begin{itemize}
  \item[(4)]
    The hydraulic jumps with random perturbations (Section 5.4.3). This is good start for introducing a
    'disturbance' but may not be sufficient to justify carbuncle-free fix. Please try out the following
    disturbance. Let $(u_L, h_L)$ to be the initial left and $(u_R, h_R)$ the right states of the jump. 
    Introduce the following disturbance at the last cell prior to the jump such that 
    $u=\delta { u_L} + (1-\delta)u_R, h=\delta {h_L} + (1-\delta)$ where $0\leq{\delta}\leq{1}$. 
    Please try a discrete increment such that $\delta = 0.0, 0.1, 0.2,...,0.9, 1.0$. 
    Details can be found in Kitamura et al. (AIAA 2007, 2009). Run the proposed range of $\delta$
    for at least one of the hydraulic jump cases.
  \end{itemize}
}
\noindent
We added the requested simulation (and cited the corresponding reference) 
for the first hydraulic jump case; see Figure 10. We obtain a similar result as before; that is, 
when the Roe solver is used with any value of $\delta$, we obtain the carbuncle instability. 
With the blended solver, we do not obtain carbuncles for any value of $\delta$.

@@

    \bibitem{ismail2009v2}
      F. Ismail, P.L. Roe, and H. Nishikawa,
      A proposed cure to the carbuncle phenomenon.
      {\it In Computational Fluid Dynamics}, (2006), 149--154.
\end{thebibliography}

%\bibliographystyle{plain}
%\bibliography{refs}
%\clearpage
%\bibliographystyle{plain}
%#\bibliographystyle{abbrvnat}
%\bibliography{ref.bib}

%%----------------------------------------------------------%%
%%-------------------- SECOND REVIEWER -------------------- %%
%%----------------------------------------------------------%%
\newpage
\section*{Revision of Manuscript \# FLD-21-0079 -- Reviewer $\#2$}
We thank the reviewer for the comments and suggestions.
We have incorporated all of them.  For your convenience,
the modifications related to the comments made by reviewer $\#1$ and $\#2$
appear in red and green respectively in the new version of the manuscript.
Statements in green below are comments from the reviewer.
In this letter, the comments made by the reviewer refer to the original
manuscript and our responses refer to the new manuscript.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(1)]
    When dealing with strong transcritical (or transonic, in the case of Euler equations) shocks, 
    the slowly-moving shock anomaly may appear 
    (see the thesis from Zaide \url{https://www.danielzaide.com/PDF/zaide_thesis.pdf})
    This anomaly is related to the discrete shock structure and appears when having a non-linear Hugoniot locus. 
    A spike in the momentum is observed downstream of the shock wave. Some authors have related the carbuncle to 
    the discrete shock structure and the unphysical values obtained in the cell containing the shock 
    (See Xie et al. \url{https://www.sciencedirect.com/science/article/pii/S0021999117306447?via\%3Dihub},
    Zaide and Roe \url{https://arc.aiaa.org/doi/10.2514/6.2011-3686}, etc). I recommend the authors to mention 
    this in the introduction and examine if the novel blended solver suffers from this anomaly (see comments below). 
    The performance of carbuncle-free solvers based on entropy arguments for the reduction of the slowly-moving 
    shock anomaly has not been yet assessed (to my knowledge).
  \end{itemize}
}
\noindent
We appreciate you brought these references as well as the slowly-moving anomaly to our attention.
We modified the introduction as suggested. In addition, in the new manuscript we investigate 
if our blended solver suffers from this problem. 
Based on our numercal experiments, the profile of the momentum contains the non-physical spike.
However, the location of steady shocks remains fixed; i.e., our solver does not suffer from 
the slowly-moving anomaly.

@@

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(2)]
    A numerical fix for the slowly-moving shock anomaly which avoids adding additional dissipation is proposed 
    by Zaide in \url{https://www.danielzaide.com/PDF/zaide_thesis.pdf}.
    The authors are recommended to include this reference in the introduction, 
    since it includes a very detailed review on numerical shockwave anomalies.
  \end{itemize}
}
\noindent
Thank you for bringing this reference to our attention.  It is indeed highly relevant; we have
added a citation in the introduction as well as in appropriate places later in the manuscript.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(3)]
    The proposed method is designed for the homogeneous SWE. 
    Please explain if this idea can be easily extended to the SWE with bed elevation, friction, etc. 
    Is there any limitation when dealing with source terms?
  \end{itemize}
}
\noindent
We are interested in considering source terms and guaranteeing positivity preservation but 
we believe that constitutes a new project and, as a result, we would prefer to leave that 
for future work. We have included a remark about this in the conclusions.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(4)]
    There is a paper by Navas-Montilla and Murillo 
    \url{https://www.sciencedirect.com/science/article/pii/S0021999118307496?via\%3Dihub}
    where the carbuncle and the slowly-moving shock anomaly for the SWE with bed elevation source term 
    have been examined together and some numerical fixes are proposed. This seems to be the only work 
    where source terms are considered in the study of shock anomalies for the SWE. 
    Please mention it in the introduction.
  \end{itemize}
}
\noindent
Thank you for bringing to our attention this extremely relevant work.  We have added a reference in
the introduction and have also expanded and modified our discussion of later results taking
this work into account.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(5)]
    In test case 4.1, flow over dry bed is considered but the details of the wet-dry interface 
    capturing method are omitted. Can you please mention or cite the algorithm for the wet-dry treatment?
  \end{itemize}
}
\noindent
Although we use the term ``dry bed'' in the section title, we note clearly in the text that we
use a non-zero depth $h_r=10^{-15}$.  Although we do not have a proof of positivity preservation
for our method, we do not observe violations of positivity.  Thus we avoid the
need to deal with wet-dry interfaces.  Note that we use the first-order method
(7) in order to reduce the potential for the appearance of negative depth
values.  This was already stated in the original manuscript, but we have made a revision
that emphasizes this.

@@

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(8)]
    I think the presence of the aforementioned unphysical values (spike in hu) are observed in Figure 8b (right) 
    and Figure 9 (right), as there is a ring of amplitude above (red color) for hu separating the supercritical 
    and subcritical regions. Can you include some comments about this?
  \end{itemize}
}
\noindent
We agree with the reviewer. We have included a remark about this in Section 5.4.2 and in the conclusions.
Our blended solver is a convex combination of Rusanov's and Roe's solvers,
both of which produce the spike. Consequently, the blended solver also produces the spike. 
We believe that, within our framework, the key ingredient to solve this problem is to modify
Rusanov's fluxes to eliminate the spike following for example \cite[Section 5.1]{navas2019improved}.
Afterwards, we can blend the modified Rusanov's fluxes with Roe's fluxes.
We leave this for future research but make a corresponding remark in the conclusions.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(9)]
    In figure 7c, can you also plot $hu$?
  \end{itemize}
}
\noindent
We included the requested plot. As expected, the non-physical spike is clear.
We also added a remark about this in Section 5.4.1.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(10)]
    In figure 7c, the label of the $Y$ axis is missing.
  \end{itemize}
}
\noindent
Fixed as requested.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(11)]
    In case 5.4.3, the authors mention that the observed instabilities have a physical base and are not 
    produced by the numerical method. When saying ``physical instabilities'', what do you mean? 
    The SWE is a 2D model which consider depth-averaged quantities. Note that the real physics may involve 
    a more developed hydraulic jump (which involves vertical velocities) which cannot be represented by a 
    pure discontinuity as done with the SWE. Please clarify in the text the meaning of ``physical instabilities''.
  \end{itemize}
}
\noindent
We have revised the discussion of this result to be more clear and to avoid
referring to the instability in the simulations as ``physical''.  We do refer by
way of comparison to a physical instability of the CHJ that has been known for
some time
(see e.g. [8, Figure 2]).  While we agree that the shallow water model omits
details that may be important to such an instability, we are not aware of any result that
would preclude a close connection between what is observed here and in experiments.
We believe this is an interesting result that merits further research.

### 2021-08-15 numerics_paper/version2/numerical_chj.tex

{\color{red}
  Next we perform a set of experiments similar to those of \cite[Section III B]{kitamura2009evaluation}.
  In these experiments, we perturb the initial condition in just one layer of cells, immediately behind the shock.
  We use an annular mesh (see Figure \ref{fig:mesh_chj}) with $90$ elements in
  the radial direction and $90$ elements in the angular direction.  Let $h(r)$ denote the exact solution
  of the ODE \eqref{steady}, and let $r_i$ denote the radius of a given cell center.  Then for all cells
  away from the shock (i.e. those with $r_i \notin [0.3+\Delta r]$) we use the exact solution averaged over the grid:
  \begin{align*}
    h_i = \frac{1}{\Delta r}\int_{r_i-\Delta r/2}^{r_i+\Delta r/2}h(r) dr,
    \qquad
    u_i = \frac{\beta}{h_ir_i}\left(\frac{x_i}{r_i}\right),
    \qquad
    v_i = \frac{\beta}{h_ir_i}\left(\frac{y_i}{r_i}\right),
  \end{align*}
  where $\beta=r_\jet h_\jet |u_\jet|$ and $(x_i,y_i)$ are the cartesian
  coordinates of the cell center.  For the layer of cells just behind the shock (with $0.3 \le r \le 0.3+\Delta r$),
  we set
  \begin{align*}
    h_i = \delta h_{i-1} + (1+\delta) h_{i+1},
    \qquad
    u_i = \frac{\beta}{h_ir_i}\left(\frac{x_i}{r_i}\right),
    \qquad
    v_i = \frac{\beta}{h_ir_i}\left(\frac{y_i}{r_i}\right),
  \end{align*}
  where $\delta=\{0,0.1,0.2,\dots,0.9,1\}$ and $h_L$ and $h_R$ are the values of the water height
  at the left and right cells from the cell with $r_i=r_s+\Delta r/2$, respectively.
  In Figure \ref{stab_shock_init_cond}, we plot representative results of these experiments, showing
  the initial data in the first column and solutions at $t=3$ in the second and third columns.
  we show only the results using $\delta=\{0,0.5,1\}$ since other results were similar.
  The solution obtained with the Roe solver, in which the carbuncle instability appears,
  is sensitive to the perturbation just described, while the solution obtained with the
  blended solver is radially symmetric and essentially the same for all values of $\delta$.
  \begin{figure}
    \subfloat[Initial condition ($t=0$)\label{stab_shock_init_cond}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0.png} 
    \end{tabular}
    }
    ~
    \subfloat[Roe's solver ($t=3$)\label{stab_shock_roe}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0_roe.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5_roe.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0_roe.png} 
    \end{tabular}
    }
    ~
    \subfloat[Blended solver ($t=3$)\label{stab_shock_blended}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0_blended.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5_blended.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0_blended.png} 
    \end{tabular}
  }    
    \caption{Simulation of a CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r1},
      using the 2nd-order method \eqref{second-order_via_fluct} with two Riemann solvers. The depth $h$
      is shown as a function of $r$.  Note that the plots in the third column contain include lines for
      each angle along the grid, but only a single line is visible due to the solution's symmetry.
      \label{fig:stab_shock}}
  \end{figure}
}

>>>>

{\color{red}
  Next we perform a set of experiments similar to those of \cite[Section III B]{kitamura2009evaluation}.
  In these experiments, we perturb the initial condition in just one layer of cells, immediately behind the shock.
  We use an annular mesh (see Figure \ref{fig:mesh_chj}) with $90$ elements in
  the radial direction and $90$ elements in the angular direction.  Let $h(r)$ denote the exact solution
  of the ODE \eqref{steady}, and let $r_i$ denote the radius of a given cell center.  Then for all cells
  away from the shock (i.e. those with $r_i \notin [0.3+\Delta r]$) we use the exact solution averaged over the grid:
  \begin{align*}
    h_i = \frac{1}{\Delta r}\int_{r_i-\Delta r/2}^{r_i+\Delta r/2}h(r) dr,
    \qquad
    u_i = \frac{\beta}{h_ir_i}\left(\frac{x_i}{r_i}\right),
    \qquad
    v_i = \frac{\beta}{h_ir_i}\left(\frac{y_i}{r_i}\right),
  \end{align*}
  where $\beta=r_\jet h_\jet |u_\jet|$ and $(x_i,y_i)$ are the cartesian
  coordinates of the cell center.  For the layer of cells just behind the shock (with $0.3 \le r \le 0.3+\Delta r$),
  we set
  \begin{align*}
    h_i = \delta h_{L} + (1+\delta) h_{R},
    \qquad
    u_i = \frac{\beta}{h_ir_i}\left(\frac{x_i}{r_i}\right),
    \qquad
    v_i = \frac{\beta}{h_ir_i}\left(\frac{y_i}{r_i}\right),
  \end{align*}
  where $\delta=\{0,0.1,0.2,\dots,0.9,1\}$ and $h_L$ and $h_R$ are the values of the water height
  at the left and right cells from the cell with $r_i=r_s+\Delta r/2$, respectively.
  In Figure \ref{stab_shock_init_cond}, we plot representative results of these experiments, showing
  the initial data in the first column and solutions at $t=3$ in the second and third columns.
  We show only the results using $\delta=\{0,0.5,1\}$ since other results were similar.
  The solution obtained with the Roe solver, in which the carbuncle instability appears,
  is sensitive to the perturbation just described, while the solution obtained with the
  blended solver is radially symmetric and essentially the same for all values of $\delta$.
  \begin{figure}
    \subfloat[Initial condition ($t=0$)\label{stab_shock_init_cond}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0.png} 
    \end{tabular}
    }
    ~
    \subfloat[Roe's solver ($t=3$)\label{stab_shock_roe}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0_roe.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5_roe.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0_roe.png} 
    \end{tabular}
    }
    ~
    \subfloat[Blended solver ($t=3$)\label{stab_shock_blended}]{
    \begin{tabular}{c}
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p0_blended.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta0p5_blended.png} \\ \\
      \includegraphics[scale=0.27,valign=c]{figures/chj_slices_delta1p0_blended.png} 
    \end{tabular}
  }    
    \caption{Simulation of a CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r1},
      using the 2nd-order method \eqref{second-order_via_fluct} with two Riemann solvers. The depth $h$
      is shown as a function of $r$.  Note that the plots in the third column include lines for
      each angle along the grid, but only a single line is visible due to the solution's symmetry.
      \label{fig:stab_shock}}
  \end{figure}
}
